Manuscript for foodborne-outbreak Project

### 2023-03-17 code/processing_code/processingfile_v1.qmd
update2

## Testing Data
```{r}
str(analysis_data)
summary(analysis_data)

# Closer Look at Training Data
head(analysis_data)
skim(analysis_data)

# Plotting NAs of Training Data
plot_na_pareto(analysis_data)
```

@@

Because the analysis dataset contains more than just foodborne outbreaks, we want to specify it to just foodborne illnesses.

```{r}
# Plot of NAs After Limiting to Analysis Data to Food
plot_na_pareto(analysis_data %>% filter(`Primary Mode` == "Food"))
```

@@

# Analysis Data
unique(analysis_data$`Etiology Status`)
```

@@

## Analysis Data
```{r}
analysis_data_fin <- analysis_data %>%
  select(!c(`Animal Type`, `Animal Type Specify`, 
            `Water Exposure`, `Water Type`, `Water Status`, 
            `Food Contaminated Ingredient`)) %>%
  filter(grepl("Confirmed", `Etiology Status`)) %>%
  filter(`Primary Mode` == "Food")
```

### Separate Rows to Address Coinfections
```{r}
analysis_data_fin <- analysis_data_fin %>%
  separate_rows(Etiology, sep = "; ")
```

@@

analysis_data_fin <- analysis_data_fin %>%
  mutate(`IFSAC Category` = case_when(
    grepl("Salad", `Food Vehicle`, ignore.case = TRUE) ~ "Salad",
    grepl(paste(Desserts, 
                collapse = "|"), `Food Vehicle`) ~ "Dessert",
    grepl("Chicken", `Food Vehicle`, ignore.case = TRUE) ~ "Poultry",
    grepl("Turkey", `Food Vehicle`, ignore.case = TRUE) ~ "Poultry",
    grepl("Poultry", `Food Vehicle`, ignore.case = TRUE) ~ "Poultry",
    TRUE ~ `IFSAC Category`))
```

@@

analysis_data_fin <- analysis_data_fin %>%
  mutate(Pathogen_Type = case_when(
    grepl(paste(Pathogen_Poison, 
                collapse = "|"), Etiology) ~ "Poison",
    grepl(paste(Pathogen_Bacteria, 
                collapse = "|"), Etiology) ~ "Bacteria",
    grepl(paste(Pathogen_Parasite, 
                collapse = "|"), Etiology) ~ "Parasite",
    grepl(paste(Pathogen_Virus, 
                collapse = "|"), Etiology) ~ "Virus"
    )) %>%
  mutate(Simplified_Etiology = word(Etiology,1)) %>%
      mutate(Simplified_Etiology = case_when(
        Simplified_Etiology == "Escherichia" ~ "E. coli",
        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
        TRUE ~ Simplified_Etiology))
```
###Column for Location
```{r}
analysis_data_fin_re <- analysis_data_fin %>%
  rename(Location=Setting)%>%
  mutate(Location=case_when(
    grepl("Restaurant",Location)~"Restaurant",TRUE~Location))%>%
    separate_rows(Location,sep="; ")
analysis_data_fin_re
```

# Drop Rows Containing NA
```{r}
training_data_final=training_data_fin%>%
  drop_na()
analysis_data_final=analysis_data_fin_re%>%
  drop_na()

@@

## Check Data Structure Again
```{r}
skim(training_data_final)
skim(analysis_data_final)
```

@@

```{r}
# Label Final Datasets as Cleaned/Processed
processed_training_data <- training_data_final
processed_analysis_data <- analysis_data_final

@@

save_data_location2=here::here("data","processed_data","processed_analysis_data.rds")
saveRDS(processed_analysis_data, file = save_data_location2)

>>>>

## Testing Data
```{r}
str(testing_data)
summary(testing_data)

# Closer Look at Testing Data
head(testing_data)
skim(testing_data)

# Plotting NAs of Training Data
plot_na_pareto(testing_data)
```

@@

Because the testing dataset contains more than just foodborne outbreaks, we want to specify it to just foodborne illnesses.

```{r}
# Plot of NAs After Limiting to Analysis Data to Food
plot_na_pareto(testing_data %>% filter(`Primary Mode` == "Food"))
```

@@

# Analysis Data
unique(testing_data$`Etiology Status`)
```

@@

## Analysis Data
```{r}
testing_data_fin <- testing_data %>%
  select(!c(`Animal Type`, `Animal Type Specify`, 
            `Water Exposure`, `Water Type`, `Water Status`, 
            `Food Contaminated Ingredient`)) %>%
  filter(grepl("Confirmed", `Etiology Status`)) %>%
  filter(`Primary Mode` == "Food")
```

### Separate Rows to Address Coinfections
```{r}
testing_data_fin <- testing_data_fin %>%
  separate_rows(Etiology, sep = "; ")
```

@@

testing_data_fin <- testing_data_fin %>%
  mutate(`IFSAC Category` = case_when(
    grepl("Salad", `Food Vehicle`, ignore.case = TRUE) ~ "Salad",
    grepl(paste(Desserts, 
                collapse = "|"), `Food Vehicle`) ~ "Dessert",
    grepl("Chicken", `Food Vehicle`, ignore.case = TRUE) ~ "Poultry",
    grepl("Turkey", `Food Vehicle`, ignore.case = TRUE) ~ "Poultry",
    grepl("Poultry", `Food Vehicle`, ignore.case = TRUE) ~ "Poultry",
    TRUE ~ `IFSAC Category`))
```

@@

testing_data_fin <- testing_data_fin %>%
  mutate(Pathogen_Type = case_when(
    grepl(paste(Pathogen_Poison, 
                collapse = "|"), Etiology) ~ "Poison",
    grepl(paste(Pathogen_Bacteria, 
                collapse = "|"), Etiology) ~ "Bacteria",
    grepl(paste(Pathogen_Parasite, 
                collapse = "|"), Etiology) ~ "Parasite",
    grepl(paste(Pathogen_Virus, 
                collapse = "|"), Etiology) ~ "Virus"
    )) %>%
  mutate(Simplified_Etiology = word(Etiology,1)) %>%
      mutate(Simplified_Etiology = case_when(
        Simplified_Etiology == "Escherichia" ~ "E. coli",
        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
        TRUE ~ Simplified_Etiology))
```
###Column for Location
```{r}
testing_data_fin <- analysis_data_fin %>%
  rename(Location=Setting)%>%
  mutate(Location=case_when(
    grepl("Restaurant",Location)~"Restaurant",TRUE~Location))%>%
    separate_rows(Location,sep="; ")
testing_data_fin
```

# Drop Rows Containing NA
```{r}
training_data_final=training_data_fin%>%
  drop_na()
testing_data_final=testing_data_fin%>%
  drop_na()

@@

## Check Data Structure Again
```{r}
skim(training_data_final)
skim(testing_data_final)
```

@@

```{r}
# Label Final Datasets as Cleaned/Processed
processed_training_data <- training_data_final
processed_testing_data <- testing_data_final

@@

save_data_location2=here::here("data","processed_data","processed_testing_data.rds")
saveRDS(processed_testing_data, file = save_data_location2)

### 2023-03-17 products/manuscript/Manuscript.qmd
update2

We adopted two datasets for this study.
The first dataset was uploaded onto Kaggle in 2017 and contains data from the Centers for Disease Control and Prevention (CDC) on foodborne disease outbreaks from 1998 to 2015 [@cdc2017].
There are 12 variables and 16,950+ observations.

The second dataset was collected from the National Outbreak Reporting System (NORS), Centers for Disease Control and Prevention (CDC) website [@cdc2020].
This dataset covers the reports of foodborne and waterborne disease outbreaks and enteric (intestinal) disease outbreaks spread by contact with environmental sources, infected people or animals, and other means from 2009 to 2020.
The raw datasets may be viewed within the repository of this project (data/raw_data/).

@@

Afterwards, we built regression models using these predictors and health outcomes; we intend to implement machine learning to train a model to predict health outcomes using the Kaggle dataset for the CDC NORS data.

>>>>

We adopted two datasets for this study.
The first dataset was uploaded onto Kaggle in 2017 and contains data from the Centers for Disease Control and Prevention (CDC) on foodborne disease outbreaks from 1998 to 2015 [@cdc2017].
There are 12 variables and 16,950+ observations.
This data set is intended to be used for model building.

The second dataset was collected from the National Outbreak Reporting System (NORS), Centers for Disease Control and Prevention (CDC) website [@cdc2020].
This dataset covers the reports of foodborne and waterborne disease outbreaks and enteric (intestinal) disease outbreaks spread by contact with environmental sources, infected people or animals, and other means from 2009 to 2020.
The raw datasets may be viewed within the repository of this project (data/raw_data/).
This data set is collected with the purpose of evaluating the model performance in the end.

@@

Afterwards, we built regression models including linear and Poisson regression using these predictors and health outcomes from training data set [@cdc2017]; data was re-sampled using bootstrapping approach for 1000 times for a better estimate of the accuracy and variability of the model coefficients; we intend to implement random forest to train a model to predict health outcomes using the Kaggle dataset for the CDC NORS data. The performance of models was compared across different metrics. The best model was ultimately employed to predict the outcome based on predictors in the testing data set [@cdc2020a].
